Geometric variational problems

### 2020-11-05 NSF2020/Description/specificaim3_ver2.tex
changes throughout aim 3

\subsection{Specific Aim 3: Efficient high-order numerical algorithm for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
\subsubsection{Overview of the potential theory and boundary integral
equation methods} 
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces an additional
source of error, and discretizing the volume, particularly in three
dimensions, results in an excessively large linear system. Moreover, it
is difficult to obtain high-order discretizations when the boundary
surfaces are irregular. Since we are solving constant-coefficient
elliptic PDEs, we apply boundary integral equation (BIE) that remove the
aforementioned obstacles. BIEs represent the solution as a layer
potential with an unknown density function. Then, the correct density
function is found that matching the layer potential with the prescribed
boundary conditions. By construction, layer potentials satisfy the
underlying PDE and the far-field conditions. For the exterior Dirichlet
problem of the screened Laplace equation, the BIE formulation in our
previous work~\cite{Fu2018_SIAM} can be readily extended to three
dimensions. For the mobility problem of the Stokes flow in three
dimensions that takes the hydrodynamic effect into account, an
well-conditioned BIE is found in~\cite{manasthesis}.

Discretizations of BIEs have several advantages when compared to
stencil-based discretizations. First, the dimension of the problem is
reduced by one since the unknown density function is defined on the
boundary of the geometry. Second, by carefully selecting each layer
potential, the resulting linear system is well-conditioned for any
resolution and iterative methods such as GMRES require a
mesh-independent number of iterations to converge. Third, by using
appropriate quadrature methods, the method converges with spectral or
very high-order accuracy. The most significant challenges include
solving dense linear systems, developing preconditioning strategies, and
developing quadrature methods for integrands that are nearly singular.

By choosing BIEs that use a mesh-independent number of GMRES iterations,
the overall cost of a solver is proportional to the complexity of the
matrix-vector multiplication. The matrix-vector multiplication will be
accelerated to optimal or near-optimal complexity using the fast
multipole method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2,
fmm3, fmm4, fmm6, fmm7, fmm8}. Alternatively, fast direct solvers can be
used~\cite{fds1, fds2, fds3, fds4, fds5, fds6, fds7, fds8, ho2016cpam2,
ho2016cpam1, minden2016, minden2017siammms}, but since direct solvers
often require a large amount of initial overhead, they are less
applicable to problems with moving geometries such as ours.
Alternatively, fast direct solvers can be used to precondition the
linear system as described in Section~\ref{subsec:NumericalIssues}.

% ----------------------------------------------------------------------
\subsubsection{Nearly touching bodies}
Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and such near-contact results in nearly-singular integrands
in both two and three dimensions. Quadrature methods to address such
integrands has received a lot of attention in two and three
dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2, davis_1984,
graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003, khayat_2005,
schwab_1992, ying_2006, beale1, beale2, goodman_1990, haroldson_1998,
lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3, helsing_2008a,
helsing_integral_2009, helsing_tutorial_2012, klockner2013jcp, qbx2,
wala2019jcp, af2018sisc, siegel2018jcp, rachh2017jcp, ding2019arxiv,
bar2014}. For two-dimensional problems, the trapezoid rule is the
workhorse for BIEs since it achieves spectral accuracy when integrands
are not nearly-singular~\cite{tre-wei2014}. To address nearly-singular
integrands of two dimensional BIEs, we will use a {\em barycentric
quadrature rule} that only requires a slight modification of the
trapezoid rule~\cite{ioa-pap-per1991}. In its current form, this method
requires the layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature rule so that it can be applied to the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as
$K(\mathbf{x},\mathbf{y}) = K_1(\mathbf{x},\mathbf{y}) +
K_2(\mathbf{x},\mathbf{y})$, where $K_1(\mathbf{x},\mathbf{y}) =
-\log\|\mathbf{x} - \mathbf{y}\|$ and $K_2(\mathbf{x},\mathbf{y}) =
K(\mathbf{x},\mathbf{y}) + \log\|\mathbf{x} - \mathbf{y}\|$. Using this
decomposition, the layer potential involving $K_1(\mathbf{x}\mathbf{y})$
can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and the layer potential
$K_2(\mathbf{x},\mathbf{y})$ can be accurately computed with the
trapezoid rule since the kernel is bounded for all
$\mathbf{x}$ and $\mathbf{y}$.

@@

% ----------------------------------------------------------------------
\subsubsection{Other numerical issues}
\label{subsec:NumericalIssues}
Preconditioning
Adaptive high-order time stepping

@@

% ----------------------------------------------------------------------
\subsubsection{A hybrid method for close-to-touching Janus particles}
\textcolor{red}{
The HAP model for membranes will inevitably lead to the case where some
Janus particles will be very close to each other. Even though QBX can
handle such case accurately, it may lead to excessively large number of
unknowns in order to resolve the unknown densities at those
close-to-touching points. In~\cite{gan2016sisc}, a hybrid scheme is
developed to treat the close-to-touching spheres for the interface
problem of the Laplace equation. The key idea is to combine the analytic
image method and spectrally accurate method of moments, where the
analytic image method will capture the most singular part of the unknown
density at the close-to-touching points. With the aid of the FMM, the
scheme achieves linear complexity with much less number of unknowns on
the material interfaces. We plan to study the hybrid scheme to treat
the close-to-touching case for both the screened Laplace equation and
the mobility problem to further improve the efficiency of the numerical
algorithm for the HAP model.
}

%% ----------------------------------------------------------------------
%\subsubsection{Other numerical issues}
%Another outstanding numerical issue pertaining to simulating the
%self-assembly of amphiphilic particles (such as lipid macromolecles) in
%a viscous solvent is the collision between amphiphilic particles. The
%hydrophobic attraction potential drives the amphiphilic particles to
%move towards each other so to minimize exposure to the solvent. Under
%such attraction force the fluid is being squeezed out as particles move
%toward each other, and the amphiphilic particles can come into physical
%contact in finite time. Such particle collisions in a dense rigid body
%suspension is a great challenge and can be a bottleneck in large-scale
%simulations. In our previous work a Leonard-Jones potential is used to
%keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
%steric interaction at short ranges introduces great numerical stiffness
%that limit the time step size. 
%
%Yan {\it et al.} proposed a collision-resolution algorithm within the
%framework of boundary integral methods to resolve the non-smooth
%many-body dynamics due to collisions by formulating the particle
%collision process as a linear complementarity problem with geometric
%`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
%particle collisions based on Yan {\it et al.}'s work after the first two
%goals (see above) are achieved. A potential complication in combining
%the algorithms for particle collision into our QBX integral formulation
%is the computational cost for the QBX to refine meshes on particles in
%collision.
%
%Once the collision-resolution algorithm is implemented, we will check
%the computing performance first before we embark on incorporating
%fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
%even at the scales of the amphiphilic particle size in some cases. We
%propose to extend the high-order time-integration scheme for a
%stochastic differential equation~\cite{fu2015pre} to this system.

>>>>

\newpage

\subsection{Specific Aim 3: Efficient high-order numerical methods for
large-scale simulations}
\label{subsec:specific_aim_3}
% ----------------------------------------------------------------------
The HAP model requires solving the exterior Dirichlet problem of the
screened Laplace equation and the mobility problem of the Stokes flow at
each time step. If discretizing these equations with stencil-based
numerical methods, the computational domain must be truncated, the
volume must be discretized, and artificial boundary conditions must be
imposed. The artificial boundary conditions introduces additional error,
and discretizing the volume results in an excessively large linear
system. Moreover, it is difficult to obtain high-order discretizations
when the boundary surfaces are irregular. Since we are solving
constant-coefficient elliptic PDEs, we represent the solution of the
PDE of interest as a layer potential 
\begin{align}
  \label{eq:LP}
  u(x) = \int_{\partial\Omega} K(x,y) \sigma(y) ds_y,
\end{align}
where $\sigma$ is an unknown density function, and $K(x,y)$ is formed
with derivatives of the fundamental solution of the underlying PDE.  By
matching the boundary condition at $x_0 \in \partial\Omega$ with the
limit of equation~\eqref{eq:LP} as $x\rightarrow x_0 \in
\partial\Omega$, a BIE for the density function is formed.
Numerical methods to solve BIE has several advantages over their
PDE-based counterpart: by construction, $u$ satisfies far-field
conditions; only $\partial\Omega$ needs to be discretized; carefully
chosen kernels $K(x,y)$ result in a well-conditioned linear system that
requires a mesh-independent number of GMRES iterations to converge; and
high-order or spectral accuracy is attainable by using appropriate
quadrature methods. In our previous work~\cite{Fu2018_SIAM} we
demonstrate that BIEs are a powerful tool to simulate two-dimensional
suspensions of amphiphilic particles. In addition to improving
two-dimensional simulations, we will extend the results to three
dimensions using a standard three-dimensional BIE of the screened
Laplace equation~\cite{ying_2006} and a well-conditioned BIE of the
mobility problem~\cite{manasthesis}.

The most significant challenges of a BIE formulation include solving
dense linear systems, developing preconditioning strategies, and
developing quadrature methods for integrands that are nearly singular.
These challenges are described in Sections~\ref{subsec:NumericalIssues}.

% ----------------------------------------------------------------------
\subsubsection{Numerical issues}
\label{subsec:NumericalIssues}
\todo[inline]{Preconditioning}
\todo[inline]{Adaptive high-order time stepping}

Discretizations of carefully chosen BIEs can be solved with a
mesh-independent number of GMRES
iterations~\cite{cam-ips-kel-mey-xue1996}. Therefore, the required CPU
time is proportional to the cost of a matrix-vector multiplication that
can be done in optimal or near-optimal time with the fast multiple
method (FMM)~\cite{fmm5} and its extensions~\cite{fmm1, fmm2, fmm3,
fmm4, fmm6, fmm7, fmm8}. Alternatively, fast direct solvers can be used
to avoid any iteration~\cite{fds1, fds2, fds3, fds4, fds5, fds6, fds7,
fds8, ho2016cpam2, ho2016cpam1, minden2016, minden2017siammms}, but
direct solvers often require a large amount of initial overhead, and
this makes them less useful for problems with moving geometries. Alternatively, techniques in fast direct solvers can be used to develop
efficient preconditioning strategies.

Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and such near-contact results in nearly-singular integrands
in both two and three dimensions. Quadrature methods to address such
integrands has received a lot of attention in two and three
dimensions~\cite{alpert, kapur, sidi, duffy, bruno1, bruno2, davis_1984,
graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003, khayat_2005,
schwab_1992, ying_2006, beale1, beale2, goodman_1990, haroldson_1998,
lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3, helsing_2008a,
helsing_integral_2009, helsing_tutorial_2012, klockner2013jcp, qbx2,
wala2019jcp, af2018sisc, siegel2018jcp, rachh2017jcp, ding2019arxiv,
bar2014}. For two-dimensional problems, the trapezoid rule is the
workhorse for BIEs since it achieves spectral accuracy when integrands
are not nearly-singular~\cite{tre-wei2014}. To address nearly-singular
integrands of two dimensional BIEs, we will use a {\em barycentric
quadrature rule} that only requires a slight modification of the
trapezoid rule~\cite{ioa-pap-per1991}. In its current form, this method
requires the layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature rule so that it can be applied to the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as $K(x,y) = K_1(x,y)
+ K_2(x,y)$, where $K_1(x,y) = -\log\|x - y\|$ and $K_2(x,y) = K(x,y) +
\log\|x - y\|$. Using this decomposition, the layer potential involving
$K_1(x,y)$ can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and the layer potential $K_2(x,y)$ can be
accurately computed with the trapezoid rule since the kernel is bounded
for all $x$ and $y$.
\todo[inline]{Image here}

% ----------------------------------------------------------------------
\subsubsection{High-order adaptive time stepping}
\label{subsec:timeStepping}
\todo[inline]{Put contact in here as well}

@@

@@

%% ----------------------------------------------------------------------
%\subsubsection{A hybrid method for close-to-touching Janus particles}
%\textcolor{red}{
%The HAP model for membranes will inevitably lead to the case where some
%Janus particles will be very close to each other. Even though QBX can
%handle such case accurately, it may lead to excessively large number of
%unknowns in order to resolve the unknown densities at those
%close-to-touching points. In~\cite{gan2016sisc}, a hybrid scheme is
%developed to treat the close-to-touching spheres for the interface
%problem of the Laplace equation. The key idea is to combine the analytic
%image method and spectrally accurate method of moments, where the
%analytic image method will capture the most singular part of the unknown
%density at the close-to-touching points. With the aid of the FMM, the
%scheme achieves linear complexity with much less number of unknowns on
%the material interfaces. We plan to study the hybrid scheme to treat
%the close-to-touching case for both the screened Laplace equation and
%the mobility problem to further improve the efficiency of the numerical
%algorithm for the HAP model.
%}
%
%
%% ----------------------------------------------------------------------
%\subsubsection{Other numerical issues}
%Another outstanding numerical issue pertaining to simulating the
%self-assembly of amphiphilic particles (such as lipid macromolecles) in
%a viscous solvent is the collision between amphiphilic particles. The
%hydrophobic attraction potential drives the amphiphilic particles to
%move towards each other so to minimize exposure to the solvent. Under
%such attraction force the fluid is being squeezed out as particles move
%toward each other, and the amphiphilic particles can come into physical
%contact in finite time. Such particle collisions in a dense rigid body
%suspension is a great challenge and can be a bottleneck in large-scale
%simulations. In our previous work a Leonard-Jones potential is used to
%keep the particles from overlapping~\cite{Fu2018_SIAM}. Such steep
%steric interaction at short ranges introduces great numerical stiffness
%that limit the time step size. 
%
%Yan {\it et al.} proposed a collision-resolution algorithm within the
%framework of boundary integral methods to resolve the non-smooth
%many-body dynamics due to collisions by formulating the particle
%collision process as a linear complementarity problem with geometric
%`non-overlapping' constraint~\cite{Yan2019}. We propose to treat the
%particle collisions based on Yan {\it et al.}'s work after the first two
%goals (see above) are achieved. A potential complication in combining
%the algorithms for particle collision into our QBX integral formulation
%is the computational cost for the QBX to refine meshes on particles in
%collision.
%
%Once the collision-resolution algorithm is implemented, we will check
%the computing performance first before we embark on incorporating
%fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
%even at the scales of the amphiphilic particle size in some cases. We
%propose to extend the high-order time-integration scheme for a
%stochastic differential equation~\cite{fu2015pre} to this system.
